---
title: "Lacrosse Data Viz"
author: "Lauren Gill"
date: "3/26/2024"
output: pdf_document
---

Load in libraries
```{r}
library(tidyverse)
library(lubridate)
```

Read in data
```{r}
player_age <- read_csv("data/player_age.csv")
player_hw <- read_csv("data/player_height_weight.csv")
player_stats <- read_csv("data/player_stats.csv")
game_stats <- read_csv("data/game_stats.csv")
```

Lets do some data cleaning
First start with game_stats dataframe
```{r}
head(game_stats)

game_stats_clean <- game_stats %>%
  rename(week = "Week",
         date = "Date",
         over_time = "OT",
         attendance = "Attendance",
         score = "Score",
         diff = "Diff",
         total_goals = "Total",
         winner = "Winner",
         teams = "Final") %>%
  separate(teams, into = c('away_team', 'home_team'), sep = "@") %>%
  #seperating  the score at the end of the name in away_team
  mutate(away_score = str_extract_all(away_team, "[0-9]+")) %>%
  #seperating  the score at the end of the name in home_team
  mutate(home_score = str_extract_all(home_team, "[0-9]+")) %>%
  #get rid of scores in the colimn now 
  mutate(across(home_team,
                ~str_replace_all(., "[0-9]+", ""))) %>%
  mutate(across(away_team,
                ~str_replace_all(., "[0-9]+", ""))) %>%
  #changing overtime to 1s and 0
  mutate(ot = case_when(over_time == "OT" ~1,
                          is.na(over_time) ~ 0)) %>%
  #now getting rid of the original over_time column
  select(!c(over_time, score)) %>%
  #trim ws
  mutate(away_team = trimws(away_team),
         home_team = trimws(home_team))
```

Now cleaning player age and player hw
```{r}
head(player_age)
player_age_clean <- player_age %>%
  #i like lowercase variable names
  rename(rank = "Rank",
         name = "Player",
         team = "Team",
         birthdate = "Birthdate",
         age = "Age") %>%
  #reformating date with the lubridate package
  mutate(birthday = dmy(birthdate),
         year_born = year(birthday)) %>%
  select(!c(birthdate)) %>%
  #some entries have (R) at the end, we want to get rid of this
  mutate(across(name,
                ~str_replace_all(., "[(][R][)]", ""))) %>%
  #trimmin the white space off the ends of some of the names
  mutate(name = trimws(name))
```

```{r}
head(player_hw)

player_hw_clean <- player_hw %>%
  separate(height, c('feet', 'inches'), "'", convert = TRUE) %>% 
    #getting rid of the " at the end of the inches column
  mutate(across(inches,
                ~str_replace_all(., "[^[:alnum:]///' ]", ""))) %>%
  #converting height to a whole number
  mutate(inches = as.numeric(inches),
         height_cm = (12*feet + inches)*2.54)
```

Finally cleaning the player_stats
```{r}
head(player_stats)
```
Never mind it looks good!

Ok so now I will join a few of these data frames
First joining player_age and player_hw
```{r}
player_info_1 <- player_age_clean %>%
  full_join(player_hw_clean, by = "name")

player_info_2 <- player_info_1 %>%
  full_join(player_stats, by = "name")
```

Adding in a fun variable (sign)
First making a function
```{r}
convert_to_sign <- function(birthday) {
  month_day <- as.Date(birthday, format = "%Y-%m-%d")
  month <- lubridate::month(month_day)
  day <- lubridate::day(month_day)
  
  case_when(
    (month == 3 & day >= 21) | (month == 4 & day <= 19) ~ "Aries",
    (month == 4 & day >= 20) | (month == 5 & day <= 20) ~ "Taurus",
    (month == 5 & day >= 21) | (month == 6 & day <= 20) ~ "Gemini",
    (month == 6 & day >= 21) | (month == 7 & day <= 22) ~ "Cancer",
    (month == 7 & day >= 23) | (month == 8 & day <= 22) ~ "Leo",
    (month == 8 & day >= 23) | (month == 9 & day <= 22) ~ "Virgo",
    (month == 9 & day >= 23) | (month == 10 & day <= 22) ~ "Libra",
    (month == 10 & day >= 23) | (month == 11 & day <= 21) ~ "Scorpio",
    (month == 11 & day >= 22) | (month == 12 & day <= 21) ~ "Sagittarius",
    (month == 12 & day >= 22) | (month == 1 & day <= 19) ~ "Capricorn",
    (month == 1 & day >= 20) | (month == 2 & day <= 18) ~ "Aquarius",
    (month == 2 & day >= 19) | (month == 3 & day <= 20) ~ "Pisces",
    TRUE ~ NA_character_
  )
}
```

```{r}

str(player_info$penalty_minutes)
player_info <- player_info_2 %>%
  #using this function to make a column for astrological sign
  mutate(sign = convert_to_sign(birthday)) %>%
  #we have two columns for team, get rid of one
  select(!c(teams)) %>%
  mutate_at(vars(games_played:face_off_wins), parse_number) %>%
  #also adding bmi and converting some variables to numbers
  mutate(weight = parse_number(weight),
         age = parse_number(age),
         rank = parse_number(rank),
         bmi = weight / ((height_cm / 100) ^ 2))


ls(player_info)
```

Now I have two main datasets to work from
player_info
game_stats_clean

Lets do some exporatory data analysis
```{r}
player_info %>%
  group_by(team) %>%
  summarize(mean_age = mean(age),
            mean_height = mean(height_cm),
            mean_weight = mean(weight)) %>%
  arrange(mean_age)

```

Any trend with month born?
```{r}
player_info %>%
  mutate(month = month(birthday)) %>%
  ggplot(aes(x = as.factor(month))) +
  geom_bar(stat = "count") +
  theme_classic()

player_info %>%
  mutate(month = month(birthday)) %>%
  group_by(month) %>%
  summarize(mean_goals = mean(goals, na.rm = TRUE)) %>%
  arrange(desc(mean_goals))

```
Seems like April and May might have slightly higher numbers

How about astrological sign
```{r}
player_info %>%
  ggplot(aes(x = as.factor(sign))) +
  geom_bar(stat = "count") +
  theme_classic()

```
Ok... slightly more aeries... could test to see if this is more signficant

What about penalty minutes vs sign?
```{r}
player_info %>%
  ggplot(aes(x = reorder(sign, -penalty_minutes), y = penalty_minutes)) +
  geom_boxplot() +
  #add means to the boxplot using yellow diamonds
  stat_summary(fun.y=mean, geom="point", shape=23, size=2, color="black", fill="yellow") +
  theme_classic()

```
Not really much correlation here

Ok lets look at a whole bunch of variables with astrological sign
```{r}
player_info %>%
  group_by(sign) %>%
  summarize(mean_age = mean(age),
            mean_height = mean(height_cm),
            mean_weight = mean(weight),
            mean_penalty = mean(penalty_minutes, na.rm = TRUE),
            mean_goals = mean(goals, na.rm = TRUE),
            mean_pp = mean(pp_goals, na.rm = TRUE))

```


Oooh goals looks interesting
```{r}
player_info %>%
  ggplot(aes(x = reorder(sign, -goals), y = goals)) +
  geom_boxplot() +
  #add means to the boxplot using yellow diamonds
  stat_summary(fun.y=mean, geom="point", shape=23, size=2, color="black", fill="yellow") +
  theme_classic()


sign_goal_lm <- lm(goals ~ year_born + bmi, data = player_info)
anova1 <- aov(sign_goal_lm)
summary(anova1)

player_info %>%
  group_by(sign) %>%
  summarize(mean_goals = mean(pts, na.rm = TRUE)) %>%
  arrange(desc(by = mean_goals))

#interesting year born, height, bmi


```


Year born vs goals
```{r}
mean(player_info$height_cm, na.rm = TRUE )

player_info %>%
  ggplot(aes(x = year_born, y = goals)) +
  geom_point() +
  theme_classic()
```

Now to analze games
First need to make the variables numeric
```{r}

game_stats_clean %>%
  group_by(home_team) %>%
  summarize(mean_home_score = mean(as.numeric(home_score)),
            mean_away_score = mean(as.numeric(away_score)))

#proportion of goals
game_stats_clean %>%
  group_by(home_team) %>%
  summarize(home = mean(as.numeric(home_score)/total_goals),
            away = mean(as.numeric(away_score)/total_goals),
            diffhome_away = as.numeric(home_score) - as.numeric(away_score))%>%
  pivot_longer(!home_team, names_to = "home_away", values_to = "prop_goals") %>%
  ggplot(aes(x = home_team, y = prop_goals, fill = home_away)) +
    geom_bar(stat = "identity", position = "dodge")

#warriors specific
game_stats_clean %>%
  filter(home_team == "Vancouver Warriors") %>%
  summarize(mean_home_score = mean(as.numeric(home_score)),
            mean_away_score = mean(as.numeric(away_score)))

game_stats_clean %>%
  group_by(winner) %>%
  summarize(mean_home_score = mean(as.numeric(home_score)),
            mean_away_score = mean(as.numeric(away_score)),
            mean_diff = mean(diff))

game_stats_clean %>%
  summarize(mean_home_score = mean(as.numeric(home_score)),
            mean_away_score = mean(as.numeric(away_score)))

```

Score by attendance
```{r}
game_stats_clean %>%
  ggplot(aes(x = attendance, y = as.numeric(home_score))) +
  geom_point() +
  geom_smooth(method = "lm")

#for warriors specifically
game_stats_clean %>%
  filter(home_team == "Vancouver Warriors") %>%
  ggplot(aes(x = attendance, y = as.numeric(home_score))) +
  geom_point()

game_stats_clean %>%
  ggplot(aes(x = attendance, y = as.numeric(away_score))) +
  geom_point() +
  geom_smooth(method = "lm")

#total goals vs attendance, nothing
game_stats_clean %>%
  ggplot(aes(x = attendance, y = as.numeric(total_goals))) +
  geom_point() +
  geom_smooth(method = "lm")


#which team has the highest attendance at home games
game_stats_clean %>%
  group_by(home_team) %>%
  summarize(mean_attendance  = mean(attendance)) %>%
  arrange(desc(by = mean_attendance))

#more home wins.. statistically significnat though?? prob not (50 vs 57)
table(game_stats_clean$winner)

```
Adding a column in game_stats for day
```{r}

game_stats_clean <- game_stats_clean %>%
  mutate(date = mdy(date),
         weekday = wday(date))

#does attendance vary by weekday
#yes monday games are less attended
game_stats_clean %>%
  ggplot(aes(x = as.factor(weekday), y = attendance)) +
  geom_boxplot()

#ok so does this effect goals, perhaps!
game_stats_clean %>%
  ggplot(aes(x = as.factor(weekday), y = total_goals)) +
  geom_boxplot()
```

```{r}
game_stats_lm <- lm(ot ~ weekday+ week+ winner, 
                    data = game_stats_clean)
summary(aov(game_stats_lm))

#only significnat thing is ot?? usually no ot on sundays
game_stats_clean %>%
  group_by(weekday) %>%
  summarize(prop_ot  = mean(ot),
            number = length(ot)) %>%
  arrange(by = prop_ot) %>%
  ggplot(aes(x= as.factor(weekday), y= prop_ot)) +
  geom_bar(stat = "identity")


game_stats_clean %>%
  group_by(weekday) %>%
  summarize(prop_ot  = mean(ot),
            number = length(ot)) %>%
  arrange(by = prop_ot)
```


## Interesting trends
- Goals by star sign (check)
- Overtime on weekdays (check)
- Home vs away games
- Best attended games (check)
- penalty minutes by height/age and blocked shots


Maybe we can look at some warrior specific stats
```{r}
player_info %>%
  filter(team == "Vancouver Warriors") %>%
  group_by(sign) %>%
  summarize(n = n()) %>%
  arrange(by = n)
#Libra and Taurus

player_info %>%
  filter(team == "Vancouver Warriors") %>%
  arrange(desc(by = penalty_minutes))

player_lm <- lm(penalty_minutes ~ age + rank + height_cm + weight + goals + loose_ball + blocked_shots + bmi, data = player_info)
summary(aov(player_lm))
#loose ball, blocked shot, age

player_info %>%
  ggplot(aes(x = age, y = penalty_minutes)) +
  geom_point(colour = "black") +
  geom_smooth(method = "lm", colour = "red", fill = "red") +
  theme_classic() +
  labs(x = "Age", y = "Penalty Minutes")

player_info %>%
  group_by(age) %>%
  summarize(mean_min = mean(penalty_minutes, na.rm = TRUE))


player_info %>%
  ggplot(aes(x = blocked_shots, y = penalty_minutes)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "red", fill = "red")

player_info %>%
  group_by(name) %>%
  summarize(mean_min = mean(penalty_minutes, na.rm = TRUE),
            mean_blocked = mean(blocked_shots, na.rm = TRUE))

player_info %>%
  ggplot(aes(x = loose_ball, y = penalty_minutes)) +
  geom_point() +
  geom_smooth(method = "lm")
# one huge outlier that seems to be driving this trend

player_info %>%
  ggplot(aes(x = height_cm, y = penalty_minutes)) +
  geom_point() +
  geom_smooth(method = "lm")

```


Height of lacrosse players
```{r}
mean(player_info$height_cm, na.rm = TRUE)
#185.27 cm

max(player_info$height_cm, na.rm = TRUE)
#203.2 cm
min(player_info$height_cm, na.rm = TRUE)
#167.64 cm

mean(player_info$age, na.rm = TRUE)
#27.8

player_info %>%
  group_by(sign) %>%
  summarize(n = n())
```

do shorter names do better/worse
```{r}
player_name <- player_info %>%
  separate(name, into =c("first", "last" ), sep = " ") %>%
  mutate(first_letters = nchar(first),
         last_letters = nchar(last))

#any player differenes, nno
player_name %>%
  ggplot(aes(x = last_letters, y = games_played)) +
  geom_point() +
  geom_smooth(method = "lm")

#by team
player_name %>%
  group_by(team) %>%
  summarize(mean_last = mean(last_letters)) %>%
  arrange(desc(mean_last))
  
mean(game_stats_clean$total_goals)
```


Do players with an animal mascot play better
```{r}
table(player_info$team)

animal <- c("Albany FireWolves", 
                  "Halifax Thunderbirds",
                  "Las Vegas Desert Dogs",
                  "Panther City Lacrosse Club",
                  "Rochester Knighthawks",
                  "San Diego Seals",
                  "Colorado Mammoth")
non_animal <- c("Buffalo Bandits", 
                "Calgary Roughnecks",
                "Georgia Swarm",
                "New York Riptide",
                "Philadelphia Wings",
                "Saskatchewan Rush",
                "Toronto Rock",
                "Vancouver Warriors")
mascot <- player_info %>%
  mutate(animal_mascot = case_when(team %in% animal ~ 1,
    team %in% non_animal ~ 0,
    TRUE ~ NA_integer_))

mascot %>%
  mutate(animal_mascot = as.factor(animal_mascot)) %>%
  group_by(animal_mascot) %>%
  summarize(mean_penalty = mean(penalty_minutes, na.rm = TRUE),
            mean_pts = mean(pts, na.rm = TRUE))

mascot_games <- game_stats_clean %>%
  mutate(animal_mascot_home = case_when(home_team %in% animal ~ 1,
    home_team %in% non_animal ~ 0,
    TRUE ~ NA_integer_))

mascot_games %>%
  mutate(animal_mascot_home = as.factor(animal_mascot_home)) %>%
  group_by(animal_mascot_home) %>%
  summarize(mean_score = mean(as.numeric(home_score), na.rm = TRUE))
```
```{r}
mean(as.numeric(game_stats_clean$home_score))
mean(as.numeric(game_stats_clean$away_score))
```

Hall of fame
```{r}
player_info %>%
  group_by(team)%>%
  summarize(mean_height = mean(height_cm),
            mean_age = mean(age),
            mean_pen = mean(penalty_minutes),
            mean_shots = mean(shots_on_goal),
            max_faceoff = mean(face_off_wins)) %>%
  arrange(desc(max_faceoff))
```

